---
title: "Data Analysis"
output: 
  html_document:
    toc: false
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(haven)
library(readxl)
library(VGAM)
```

**Regression Model**

Visit [report](report.html) for detailed steps on how the data is cleaned.
```{r}
brfss_22 = read_csv("NYSDOH_BRFSS_Survey_Data_2022.csv")|>
  janitor::clean_names()

cleaned_df = 
  brfss_22 |>
  filter(complete.cases(imprace, educag, incomg1, strsmeal_sa, genhlth))|>
  select(imprace, educag, incomg1, strsmeal_sa, genhlth)

cleaned_df = 
  cleaned_df |>
  mutate(
    income = case_when(
      incomg1 %in% c("Less than $15,000", "$25,000 to < $35,000", "$35,000 to < $50,000") ~"<50,000",
      incomg1 == "$50,000 to < $100,000" ~ "50,000 ~ 100,000",
      incomg1 %in% c ("$100,000 to < $200,000" , "$200,000 or more") ~ ">100,000",
      TRUE ~ NA_character_),
    
    education = case_when(
      educag == "Did not graduate High School" ~ "Less than high school",
      educag == "Graduated High School" ~ "High school or GED",
      educag == "Attended College or Technical School" ~ "Some post-high school",
      educag == "Graduated from College or Technical School" ~ "College graduate"
    ),
    
    food_insecure = case_when(
      strsmeal_sa %in% c("Rarely","Sometimes","Usually", "Always") ~ 1,
      strsmeal_sa == "Never" ~ 0),
    
    health = case_when(
      genhlth %in% c("Good", "Very good", "Excellent") ~ 1,
      genhlth %in% c("Fair", "Poor") ~ 0
    )
   )|>
  filter(!is.na(income), !is.na(education))|>
  select(-incomg1, -educag, -strsmeal_sa, -genhlth)

```



### Key Socioeconomic Factors Distribution

```{r}
ggplot(data = cleaned_df, aes(x = imprace, fill = imprace))+
  geom_bar() +
  labs(
    title = "Distribution of Race/Ethnicity",
    x = "Race/Ethnicity",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```



This bar graph illustrates the distribution of race/ethnicity in the dataset. The categories include American Indian/Alaskan Native, Asian, Black, Hispanic, Other Race, and White, all specified as Non-Hispanic except for the Hispanic group. Among these groups, White, Non-Hispanic individuals dominate, with the highest count exceeding 6,000. The second most prevalent group is Hispanic, followed by Black, Non-Hispanic, while Other Race, Non-Hispanic, and Asian, Non-Hispanic, show significantly smaller counts. The American Indian/Alaskan Native, Non-Hispanic category represents the smallest group. 


```{r}
ggplot(cleaned_df, aes(x = "", fill = education)) +
  geom_bar(width = 1, stat = "count") +
  coord_polar(theta = "y") +
  labs(
    title = "Distribution of Education Levels",
    fill = "Education Level"
  ) +
  theme_void() +
  theme(legend.position = "right")
```




This pie chart shows the distribution of education levels within a dataset. The largest portion of the chart is represented by college graduates, shown in red, indicating that a significant proportion of individuals have completed higher education. The second-largest category, represented in purple, corresponds to individuals with some post-high school education. High school graduates or those with a GED, shown in green, form a smaller segment, while the smallest portion of the chart, in teal, represents individuals with less than a high school education. 


```{r}
ggplot(cleaned_df, aes(x = as.factor(food_insecure), fill = as.factor(food_insecure))) +
  geom_bar() +
  scale_fill_manual(values = c("0" = "lightblue", "1" = "pink"), labels = c("Not Food Insecure", "Food Insecure")) +
  labs(
    title = "Distribution of Food Insecurity",
    x = "Food Insecurity",
    y = "Count",
    fill = "Status"
  ) +
  theme_minimal()
```




This bar graph presents the distribution of food insecurity status among individuals in the dataset. The largest group, represented in light blue, comprises individuals who are not food insecure, with a count exceeding 4,000. The second-largest group, shown in pink, represents individuals who are food insecure, numbering slightly over 2,000. A very small proportion, represented in gray, corresponds to missing data (NA) for food insecurity status. 


```{r}
cleaned_df = 
  cleaned_df |>
  mutate(
    imprace = as.factor(imprace),
    income = as.factor(income),
    education = as.factor(education)
  )
```


## Logistic Regression Models

### Is income associated with food insecure? 
```{r}
logit_income = 
  glm(formula = food_insecure ~ income, data = cleaned_df, family = binomial)

exp_income = exp(cbind(OR=coef(logit_income),confint(logit_income)))

knitr::kable(exp_income)
```

Intercept: The odds ratio for the intercept (1.171) represents the baseline odds of being food insecure when income is less than 50,000. Since the 95% confidence interval (1.079–1.272) does not include 1, it is statistically significant.

OR: 
   Income > 100,000: The odds ratio (0.166) indicates that individuals with an income greater than $100,000 have 83.4% lower odds of being food insecure compared to the income group less than 50,000. 

   Income 50,000–100,000: The odds ratio (0.349) indicates that individuals with an income between $50,000 and 100,000 have 65.1% lower odds of being food insecure compared to the income group less than 50,000. 
   
Statistical significance: confidence intervals of OR of both predictor levels exclude 1, indicating that they are statistically associated with food insecure. 



### Confidence Interval
```{r}
ci_income = as.data.frame(exp_income)
ci_income = 
  ci_income |> 
  rownames_to_column("Variable") |> 
  filter(Variable != "(Intercept)") 
```

```{r}
ggplot(ci_income, aes(y = Variable, x = OR, xmin = `2.5 %`, xmax = `97.5 %`)) +
  geom_pointrange() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Odds Ratios with 95% Confidence Intervals",
    x = "Odds Ratio (OR)",
    y = "Income Category"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12))
```



### Is race associated with food insecure?
```{r}
logit_race = 
  glm(food_insecure ~ imprace, data = cleaned_df, family = binomial)

exp_race = exp(cbind(OR=coef(logit_race),confint(logit_race)))

knitr::kable(exp_race)
```

Intercept: The odds ratio for the intercept (1.6) represents the odds of being food insecure for the reference group (American Indian/Alaskan Native). 

OR:
   Asian, Non-Hispanic: The odds ratio (0.318) indicates that Asian, Non-Hispanic individuals have 68.2% lower odds of being food insecure compared to American Indians. 

   Black, Non-Hispanic: The odds ratio (0.528) suggests that Black, Non-Hispanic individuals have 47.2% lower odds of being food insecure compared to American Indians.

   Hispanic: The odds ratio (0.832) implies that Hispanic individuals have 16.8% lower odds of being food insecure compared to American Indians. 

   Other Race, Non-Hispanic: The odds ratio (0.426) indicates that individuals of Other Race, Non-Hispanic have 57.4% lower odds of being food insecure compared to American Indians. 

   White, Non-Hispanic: The odds ratio (0.230) suggests that White, Non-Hispanic individuals have 77% lower odds of being food insecure compared to American Indians. 

Statistical significance: confidence intervals of OR of all predictor levels except Black and Hispanic exclude 1, indicating that these predictor levels are statistically significant. 



### Confidence Interval
```{r}
ci_race = as.data.frame(exp_race)
ci_race = 
  ci_race |> 
  rownames_to_column("Variable") |> 
  filter(Variable != "(Intercept)") 
```

```{r}
ggplot(ci_race, aes(y = Variable, x = OR, xmin = `2.5 %`, xmax = `97.5 %`)) +
  geom_pointrange() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Odds Ratios with 95% Confidence Intervals",
    x = "Odds Ratio (OR)",
    y = "Income Category"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12))
```





### Is education associated with food insecure?
```{r}
logit_edu = 
  glm(food_insecure ~ education, data = cleaned_df, family = binomial)

exp_edu = exp(cbind(OR=coef(logit_edu),confint(logit_edu)))

knitr::kable(exp_edu)
```


Intercept: The odds ratio for the intercept (0.309) represents the odds of being food insecure for individuals in the reference group (College graduates). 

OR:
   High School or GED: The odds ratio (2.444) indicates that individuals with a high school diploma or GED have 144.4% higher odds of being food insecure compared to college graduates. 

   Less than High School: The odds ratio (5.241) suggests that individuals with less than a high school education have 424.1% higher odds of being food insecure compared to college graduates. 

   Some Post-High School: The odds ratio (2.119) indicates that individuals with some post-high school education have 111.9% higher odds of being food insecure compared to college graduates. 

Statistical significance: confidence intervals of OR of all three predictor levels exclude 1, making these results statistically significant. 



### Confidence Interval
```{r}
ci_edu = as.data.frame(exp_edu)
ci_edu = 
  ci_edu |> 
  rownames_to_column("Variable") |> 
  filter(Variable != "(Intercept)") 
```

```{r}
ggplot(ci_edu, aes(y = Variable, x = OR, xmin = `2.5 %`, xmax = `97.5 %`)) +
  geom_pointrange() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Odds Ratios with 95% Confidence Intervals",
    x = "Odds Ratio (OR)",
    y = "Income Category"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12))
```



### Are race, education, and income associated with food insecure?
```{r}
logit_all = 
  glm(food_insecure ~ education + income + imprace, data = cleaned_df, family = binomial)

exp_all = exp(cbind(OR=coef(logit_all),confint(logit_all)))

knitr::kable(exp_all)
```

Statistical Significant Predictors: 

Education:
   Individuals with a high school diploma or GED have 40.1% higher odds of being food insecure compared to college graduates. 
   Individuals with less than a high school education have 67.1% higher odds of being food insecure compared to college graduates. 
   Individuals with some post-high school education have 49.1% higher odds of being food insecure compared to college graduates. 
    
Income: 
   Individuals earning over 100,000 have 77.2% lower odds of being food insecure compared to those earning ≤ $50,000.
   Individuals earning between $50,000 and $100,000 have 56.3% lower odds of being food insecure compared to those earning ≤ $50,000. 
   
Race/Ethnicity:
   White, Non-Hispanic individuals have 61.4% lower odds of being food insecure compared to American Indians.
   
Summary: 
   Education and income levels show significant associations with food insecurity, with lower education and income levels increasing odds of food insecurity.Among racial/ethnic groups, only White, Non-Hispanic individuals show significantly lower odds of food insecurity compared to American Indians.
   




### Confidence Interval
```{r}
ci_all = as.data.frame(exp_all)
ci_all = 
  ci_all |> 
  rownames_to_column("Variable") |> 
  filter(Variable != "(Intercept)") 
```

```{r}
ggplot(ci_all, aes(y = Variable, x = OR, xmin = `2.5 %`, xmax = `97.5 %`)) +
  geom_pointrange() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Odds Ratios with 95% Confidence Intervals",
    x = "Odds Ratio (OR)",
    y = "Income Category"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12))
```


logit_all = 
  glm(food_insecure ~ education + income + imprace, data = cleaned_df, family = binomial)

### Is food insecure associated with general health?
```{r}
logit_health = 
  glm(health ~ food_insecure, data = cleaned_df, family = binomial)

exp_health = exp(cbind(OR=coef(logit_health),confint(logit_health)))

knitr::kable(exp_health)
```

OR: 
  Individuals who are food insecure have 73.2% lower odds of reporting "Poor" health compared to "Fair" health when compared to non-food insecure individuals.
  Food-insecure individuals have 67.4% lower odds of reporting "Good" health compared to "Fair" health than non-food insecure individuals.
  Food-insecure individuals have 33.4% lower odds of reporting "Very Good" health compared to "Fair" health than non-food insecure individuals.
  Food-insecure individuals have 59.6% higher odds of reporting "Excellent" health compared to "Fair" health than non-food insecure individuals.

Statistical significance: confidence intervals of OR of "Good", "Very Good", and "Excellent" health group exclude 1, indicating that these predictor levels are statistically significant. 

This analysis emphasizes that food insecurity significantly affects perceived health levels. The result that individuals with food insecure have a higher odds ratio to report "Excellent" health compared to "Fair" health is very weird. The result may be influenced by individuals' perceptions, cultural norms, and personal experiences. Furthermore, individuals who recently are experiencing food insecurity might still maintain good physical health but are at the early stages of economic or food-related struggles, leading them to perceive their health as excellent. Therefore, the limitation of this regression is excluding confouders. 

### Confidence Interval
```{r}
coefs <- coef(summary(logit_health))
odds_ratios <- exp(coefs[, "Estimate"]) # Convert coefficients to odds ratios
lower_ci <- exp(coefs[, "Estimate"] - 1.96 * coefs[, "Std. Error"]) # Lower 95% CI
upper_ci <- exp(coefs[, "Estimate"] + 1.96 * coefs[, "Std. Error"]) # Upper 95% CI

# Combine into a tidy data frame
ci_data <- data.frame(
  Variable = rownames(coefs),
  OR = odds_ratios,
  Lower = lower_ci,
  Upper = upper_ci
) |>
  filter(!grepl("(Intercept)", Variable)) |> # Exclude intercepts
  separate(Variable, into = c("Outcome", "Predictor"), sep = ": ", fill = "right")
```


```{r}
ggplot(ci_data, aes(y = Outcome, x = OR, xmin = Lower, xmax = Upper, color = Predictor)) +
  geom_pointrange() +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Odds Ratios with 95% Confidence Intervals",
    x = "Odds Ratio (OR)",
    y = "Outcome (General Health)",
    color = "Predictor"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 12))
```

